\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage[
    backend=biber,
    style=numeric,
    sorting=none,
    maxnames=99,        % citation 中显示全部作者
    maxbibnames=99,     % bibliography 中显示全部作者
    minnames=99         % 禁止触发 et al.
]{biblatex}
\addbibresource{.bib}
\usepackage{xcolor}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

\title{Menstrual Cycle Prediction from Wearable Physiological Data:
The Role of Personalisation in Irregular Cycles\\
{\footnotesize \textsuperscript{*}Note: Sub-titles are not captured in Xplore and
should not be used}

}

\author{\IEEEauthorblockN{1\textsuperscript{st} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{2\textsuperscript{nd} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{3\textsuperscript{rd} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{4\textsuperscript{th} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{5\textsuperscript{th} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{6\textsuperscript{th} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
}

\maketitle

\begin{abstract}


( I will add it when I finish other parts.)
\end{abstract}

\begin{IEEEkeywords}
menstrual cycle prediction, wearable physiological data, personalised modelling, cycle irregularity, digital health
\end{IEEEkeywords}

\section{Introduction}
Accurate prediction of menstrual cycle events, particularly menstruation onset and ovulation timing, is important for symptom management and broader menstrual health assessment. However, the menstrual cycle is not adequately represented by a fixed-length pattern commonly idealised as 28 days \cite{bull2019realworld,grieger2020global}. Large-scale analyses of self-tracked data have shown substantial variation in cycle length, phase duration, and between-cycle consistency, and have also shown that ovulation does not reliably occur at the cycle midpoint \cite{symul2019assessment,bull2019realworld,li2020characterizing}. Traditional calendar-based approaches therefore inherit a structural limitation: they rely on historical averages and assumptions of regularity that are often least appropriate for the users who most need accurate prediction, especially users with irregular cycles. Although self-tracked menstrual records remain affected by incomplete adherence, missing observations, and heterogeneous tracking behaviour \cite{symul2019assessment,li2021predictive,symul2022labeling}, large app-based cohorts have consistently shown that cycle variability is not a marginal phenomenon but a central characteristic of menstrual health \cite{li2020characterizing,bull2019realworld}. These findings suggest that irregularity merits explicit consideration in real-world menstrual cycle prediction, rather than being treated only as residual noise around an otherwise regular pattern.

Wearable sensing is relevant here because it provides physiological information beyond calendar history. Prior studies and a recent systematic review suggest that temperature, heart rate, heart rate variability, respiratory measures, and related signals can support cycle-phase or fertile-window inference, and may outperform calendar-based or self-measured temperature approaches in some settings \cite{goodale2019wearable,maijala2019nocturnal,yu2022tracking,thigpen2025oura,luo2025prediction,kilungeja2025machine,masuda2025machine,shi2026diagnostic}. Even so, reported performance is commonly summarised by aggregate accuracy or mean error, and the underlying sources of failure are often left under-examined. Several studies note degraded performance in irregular cycles, but irregularity is usually handled as a single broad category rather than as a structured property of the user or cycle \cite{yu2022tracking,thigpen2025oura,luo2025prediction,shi2026diagnostic}. Irregularity does not represent a single condition or pattern. Across users, it may reflect different underlying clinical states, and across cycles, it may appear as consistently shifted cycle lengths, substantial cycle-to-cycle variability, or both \cite{balen2024figo,pena2018majority,bull2019realworld,li2020characterizing}. What remains unclear is whether the gains from wearable physiological signals and user-specific history are broadly shared across irregular users, or whether they are concentrated in particular irregularity patterns. Existing evaluations rarely frame this question explicitly, often reporting aggregate performance or treating irregular cycles as a single category \cite{yu2022tracking,luo2025prediction,shi2026diagnostic}. As a result, the conditions under which wearable signals and personal history materially reduce prediction error remain insufficiently understood.

Using \textit{mcPHASES}, a recently released PhysioNet dataset that combines wearable, hormonal, and self-reported menstrual health observations \cite{lin2025mcphases}, we investigate how prediction performance changes across irregularity profiles and how much benefit is obtained from incorporating user-specific history. We ask which users benefit, under what form of irregularity, and whether personalisation is uniformly useful or mainly valuable in harder cases. To address these questions, we adopt a strict subject-wise evaluation protocol and compare population-level prediction with zero-shot, one-shot, and few-shot personalised settings. More broadly, the paper reframes menstrual cycle prediction as a heterogeneity-sensitive problem and evaluates distinct forms of irregularity rather than collapsing them into a single regular-versus-irregular dichotomy.
\section{Related Work}

\subsection{Calendar-based and self-tracked cycle prediction}

Calendar-based and self-tracked approaches form an important baseline family for menstrual event prediction. Early evaluations of commercial websites and smartphone applications showed that fertile-window and ovulation estimates varied substantially across tools, and that many consumer systems relied on simplified cycle-length assumptions rather than physiological markers \cite{setton2016accuracy,moglia2016evaluation,duane2016performance}. At larger scale, app-based self-tracking records have nevertheless proved useful for characterising population-level menstrual patterns, including variation in cycle length, phase timing, and cycle-to-cycle consistency \cite{symul2019assessment,bull2019realworld,grieger2020global,li2020characterizing}. This line of work has also made clear that adherence, missingness, and reporting behaviour are not peripheral nuisances but part of the modelling problem itself \cite{li2021predictive,symul2022labeling}. Existing next-cycle prediction methods in this family therefore combine cycle-history information with adherence-aware modelling or latent-state labelling \cite{li2021predictive,symul2022labeling}. These approaches provide strong and practical baselines, but they remain largely dependent on historical and self-reported information.

\subsection{Wearable physiology for menstrual event inference}

A second line of work uses wearable sensing to access physiological signals that vary across the cycle under free-living conditions. Prior studies have examined skin or wrist temperature, heart rate, heart-rate variability, respiratory rate, and related nocturnal features for tasks including fertile-window detection, cycle-phase identification, ovulation inference, and menstruation prediction \cite{goodale2019wearable,maijala2019nocturnal,alzueta2022tracking,jasinski2024cardiovascular,yu2022tracking}. This direction is motivated in part by evidence that the fertile window is broader and less predictable than simple midpoint heuristics suggest \cite{wilcox2000fertilewindow}. Overall, wearable-based studies suggest that physiology-informed models can improve upon calendar-only or single-signal baselines in some settings \cite{goodale2019wearable,yu2022tracking}. However, this literature remains difficult to compare directly because studies differ in cohort design, reference standard, prediction window, and reported metric. Reported gains are also usually summarised through aggregate accuracy or mean error, rather than through subgroup-specific improvements or failure modes.

\subsection{Irregularity and personalisation as an evaluation gap}

The literature is more limited when it comes to how irregularity and personalisation are operationalised in evaluation. Clinically, irregular cycles do not correspond to a single physiological state: ovulatory dysfunction is one important cause, but irregular cycles are not uniformly anovulatory \cite{pena2018majority}. From a data perspective, large app-based cohorts show heterogeneity both in average cycle length and in cycle-to-cycle variability, suggesting that irregularity is better treated as a structured temporal property than as a simple binary label \cite{bull2019realworld,li2020characterizing}. Some wearable studies report lower performance in irregular users or evaluate regular and irregular groups separately \cite{yu2022tracking}, and prior work on self-tracking and fertility estimation shows that individual history can improve prediction \cite{li2016personalised,li2021predictive}. Yet these strands are rarely brought together in a single evaluation framework. In particular, there is still limited evidence on how wearable physiology and user-specific history interact with different forms of irregularity under strict subject-wise assessment. This leaves open a central question for real-world prediction: whether additional physiological and historical information benefits irregular users broadly, or mainly helps under particular irregularity patterns.

\section{Methodology}

(This section will be revised in a later stage, the current version largely follows the methodology described in the interim report.)

This study develops a wearable-based menstrual-cycle forecasting pipeline that integrates physiological signals, cycle-history information, and strict subject-wise evaluation. The methodological design follows a physiology-informed modelling perspective: rather than relying solely on calendar extrapolation, the framework aims to exploit temporal patterns in wearable physiological signals that reflect hormonal dynamics across the menstrual cycle.

\subsection{Method Overview}

The proposed framework models menstrual-cycle dynamics through a sequential analysis of physiological observations collected from wearable devices. Prior to modelling, the raw physiological signals are processed to identify ovulation-related temporal anchors and to construct aligned day-level feature sequences. Each cycle is represented as a temporally ordered sequence of daily physiological measurements and derived features.

Based on these day-level inputs, a prediction model estimates the timing of the upcoming menstrual onset. The framework supports both population-level modelling and personalised prediction settings. In the population setting, the model is trained using data from multiple individuals and applied to unseen users. In the personalised setting, limited subject-specific historical cycles are incorporated to adapt predictions to individual physiological patterns.

This design enables systematic evaluation of how prediction accuracy changes under different personalisation scenarios, particularly when only a small number of previous cycles are available.

\subsection{Dataset and Data Processing}

The study uses the mcPHASES dataset, a longitudinal wearable dataset containing multi-modal physiological signals together with hormone-based ovulation annotations. The dataset includes wrist temperature, heart rate, heart-rate variability, respiratory rate, and sleep-related measurements recorded continuously by wearable devices.

Raw wearable signals are first processed to construct daily aggregated features. Physiological measurements recorded at higher temporal resolution are summarised into nightly or daily statistics to reduce noise and ensure consistent temporal alignment across modalities. Cycles are segmented using recorded menstrual onset dates, and each cycle is represented as a sequence of daily observations.

Because wearable datasets often contain missing values due to device removal or irregular usage, the preprocessing pipeline includes quality filtering and aggregation procedures that preserve temporal continuity while minimising measurement artefacts.

\subsection{Physiological Feature Construction}

Ovulation acts as the primary physiological anchor for menstrual-cycle modelling. Several wearable signals exhibit characteristic temporal patterns around ovulation, including increases in peripheral skin temperature and changes in heart-rate variability. To capture these patterns, the feature-construction process combines cycle-history information with day-level physiological measurements.

Temperature-related features are derived from wrist temperature trajectories, which often display a biphasic pattern across ovulatory cycles. Heart-rate and HRV features capture autonomic nervous system responses associated with hormonal fluctuations. Respiratory rate and sleep metrics are incorporated as additional indicators of physiological state.

To account for substantial inter-individual variability in physiological baselines, features are normalised using subject-specific rolling statistics. For a given feature value $x_t$ at day $t$, normalisation is performed relative to the mean and standard deviation of preceding observations for that individual. This procedure emphasises deviations from personal baseline patterns rather than absolute physiological levels.

\subsection{Prediction Models and Personalisation}

Prediction models are trained to estimate the timing of the next menstrual onset based on the constructed feature sequences. In this study, emphasis is placed on analysing the effect of personalisation rather than introducing highly complex architectures. Therefore, the primary models used in experiments are relatively interpretable regression-based learners.

Two prediction settings are considered. In the population setting, the model is trained using pooled data from multiple users and evaluated on unseen individuals. In the personalised setting, limited subject-specific history is incorporated. We analyse several personalisation regimes, including:

\begin{itemize}
\item \textbf{0-shot prediction}: no historical cycles from the target user are available.
\item \textbf{1-shot personalisation}: the model has access to one previous cycle.
\item \textbf{few-shot personalisation}: two or more previous cycles are available.
\end{itemize}

These regimes allow investigation of how prediction accuracy improves as additional subject-specific information becomes available.

\subsection{Evaluation Protocol}

Evaluation is performed using strict subject-wise data splits in order to simulate realistic deployment scenarios. Under this protocol, cycles from the same individual are not simultaneously used for both training and testing in the population setting.

Prediction performance is assessed using standard error metrics for cycle-timing estimation, including mean absolute error (MAE) and window-based accuracy (e.g., predictions within $\pm3$ days of the true onset date). In addition to overall performance, results are stratified by cycle irregularity to examine whether population models degrade as irregularity increases and whether personalised models yield greater improvements for users with irregular cycles.

This evaluation framework enables systematic analysis of the interaction between cycle irregularity and model personalisation under realistic data constraints.
\section{Experiments and Results}

This section evaluates menstrual onset prediction performance under subject-level evaluation and analyses how prediction accuracy varies with cycle irregularity and model personalisation.

\subsection{Dataset}

Experiments are conducted using the mcPHASES dataset, a longitudinal wearable dataset containing multi-modal physiological measurements together with hormone-based ovulation annotations. The dataset includes wrist temperature, heart rate, heart-rate variability (HRV), respiratory rate, and sleep metrics collected continuously by wearable devices.

Following preprocessing, each menstrual cycle is represented as a sequence of daily physiological observations and derived features. Cycles with insufficient physiological data coverage are excluded during preprocessing to ensure consistent signal quality. After filtering, the dataset contains [X] participants and [X] menstrual cycles.

For each cycle, the prediction target is the onset date of the next menstruation. All features used in prediction are restricted to information available at prediction time, ensuring that no future-cycle information or hormone measurements are used as model inputs.

\subsection{Experimental Settings}

To evaluate generalisation performance, we adopt a strict subject-wise evaluation protocol. In this setting, users appearing in the test set are not included in the training set for population models. This design simulates real-world deployment where predictions must generalise to previously unseen individuals.

Two modelling scenarios are considered:

\textbf{Population model.}  
A single model is trained on pooled data from multiple users and evaluated on unseen individuals.

\textbf{Personalised model.}  
Subject-specific historical cycles are incorporated when predicting future cycles for the same user. We analyse several personalisation regimes:

\begin{itemize}
\item 0-shot prediction (no historical cycles from the target user)
\item 1-shot personalisation (one historical cycle available)
\item 2-shot personalisation (two historical cycles available)
\end{itemize}

These settings allow analysis of how prediction accuracy changes as additional user-specific data becomes available.

\subsection{Cycle Irregularity Definition}

To investigate the effect of cycle variability, users are stratified according to menstrual cycle irregularity. For each participant, cycle irregularity is quantified using the standard deviation of cycle length across observed cycles.

Based on this measure, users are divided into two groups:

\begin{itemize}
\item \textbf{Regular cycles:} low variability in cycle length
\item \textbf{Irregular cycles:} high variability in cycle length
\end{itemize}

This grouping enables comparison of prediction performance between relatively stable and highly variable menstrual cycles.

\subsection{Evaluation Metrics}

Prediction accuracy is evaluated using two metrics commonly used in menstrual-cycle forecasting:

\begin{itemize}
\item \textbf{Mean Absolute Error (MAE):} the absolute difference between the predicted onset day and the true onset day.
\item \textbf{$\pm3$ day accuracy:} the proportion of predictions that fall within three days of the true onset date.
\end{itemize}

These metrics capture both average prediction error and clinically meaningful timing accuracy.

\subsection{Population Model Performance}

We first evaluate the performance of the population model under subject-wise evaluation. Table~I reports prediction accuracy across all users.

\begin{table}[h]
\centering
\caption{Population model performance}
\begin{tabular}{ccc}
\hline
Model & MAE (days) & $\pm3$d Accuracy \\
\hline
Population Model & [X.X] & [XX]\% \\
\hline
\end{tabular}
\end{table}

To examine the impact of cycle irregularity, we further analyse performance separately for regular and irregular users. Results are summarised in Table~II.

\begin{table}[h]
\centering
\caption{Population model performance by cycle regularity}
\begin{tabular}{ccc}
\hline
User Group & MAE (days) & $\pm3$d Accuracy \\
\hline
Regular cycles & [X.X] & [XX]\% \\
Irregular cycles & [X.X] & [XX]\% \\
\hline
\end{tabular}
\end{table}

These results suggest that prediction errors increase when cycle irregularity is high, indicating that population-level models struggle to capture the variability present in irregular menstrual cycles.

\subsection{Effect of Personalisation}

We next analyse how prediction accuracy changes when limited subject-specific history is available. Table~III compares prediction performance under different personalisation regimes.

\begin{table}[h]
\centering
\caption{Effect of personalisation}
\begin{tabular}{ccc}
\hline
Setting & MAE (days) & $\pm3$d Accuracy \\
\hline
0-shot & [X.X] & [XX]\% \\
1-shot & [X.X] & [XX]\% \\
2-shot & [X.X] & [XX]\% \\
\hline
\end{tabular}
\end{table}

Overall, incorporating a small number of historical cycles improves prediction accuracy compared with the zero-shot setting.

\subsection{Personalisation Benefits for Irregular Cycles}

Finally, we examine whether personalisation provides larger improvements for users with irregular cycles. Table~IV reports the change in prediction error between the 0-shot and personalised settings for both user groups.

\begin{table}[h]
\centering
\caption{Personalisation benefit by cycle regularity}
\begin{tabular}{ccc}
\hline
User Group & MAE Reduction (days) & Accuracy Gain \\
\hline
Regular cycles & [X.X] & [XX]\% \\
Irregular cycles & [X.X] & [XX]\% \\
\hline
\end{tabular}
\end{table}

Preliminary results suggest that personalisation yields larger improvements for irregular users, indicating that subject-specific physiological patterns may help explain cycle variability that population models fail to capture.

\section{Discussion}

The experimental results provide several insights into wearable-based menstrual-cycle prediction and the role of personalisation in modelling physiological time series.

First, the results suggest that prediction performance degrades as cycle irregularity increases. Population models trained across multiple users perform reasonably well for individuals with relatively stable cycle patterns but show larger prediction errors for users with high cycle-length variability. This observation is consistent with physiological knowledge: while the luteal phase tends to be relatively stable, variability in ovulation timing introduces substantial uncertainty in the follicular phase, which directly affects the timing of the next menstrual onset.

Second, the analysis indicates that incorporating subject-specific historical cycles improves prediction accuracy. Even limited personalisation, such as access to one or two previous cycles, reduces prediction error compared with the zero-shot population setting. This finding suggests that individual physiological patterns captured by wearable signals contain useful information that population-level models cannot fully exploit.

More importantly, the results suggest that the benefit of personalisation may not be uniform across users. Preliminary analyses indicate that personalisation provides larger improvements for users with irregular cycles than for users with regular cycles. This observation supports the hypothesis that cycle irregularity may reflect individual-specific physiological dynamics rather than purely random noise. Personalised models may therefore help capture these individual patterns when sufficient historical data becomes available.

These findings have broader implications for digital health systems that rely on wearable physiological data. Rather than focusing solely on improving overall prediction accuracy, it may be equally important to understand which user groups benefit most from personalisation and how quickly models can adapt to new users.

Several limitations should also be noted. First, the dataset used in this study contains a limited number of participants, which constrains the statistical power of subgroup analyses. Second, wearable physiological signals are inherently noisy and may be affected by behavioural and environmental factors unrelated to menstrual physiology. Finally, the current models focus primarily on prediction accuracy and do not explicitly model the biological mechanisms underlying hormonal regulation.

Future work may therefore explore larger datasets, more advanced sequence models, and improved representations of physiological dynamics. In particular, modelling approaches that explicitly represent latent cycle states or hormonal transitions may provide a more principled way to integrate wearable signals with reproductive physiology.

\section{Conclusion}

This study investigates menstrual onset prediction using wearable physiological data, with a particular focus on the interaction between cycle irregularity and model personalisation.

Using a subject-wise evaluation protocol on the mcPHASES dataset, we analyse prediction performance under both population and personalised modelling settings. The results suggest that population models perform less reliably for individuals with highly irregular cycles. Incorporating subject-specific historical cycles improves prediction accuracy, and preliminary evidence indicates that the benefit of personalisation may be greater for users with irregular menstrual patterns.

These findings highlight the importance of considering user heterogeneity when designing wearable-based reproductive health prediction systems. Rather than assuming uniform model performance across users, future research should examine how prediction accuracy varies across different physiological profiles and how quickly personalised models can adapt to individual users.

Further work will extend this study by exploring additional modelling approaches, larger datasets, and more detailed analyses of physiological signal dynamics across the menstrual cycle.



\printbibliography
\vspace{12pt}
\color{red}
IEEE conference templates contain guidance text for composing and formatting conference papers. Please ensure that all template text is removed from your conference paper prior to submission to the conference. Failure to remove the template text from your paper may result in your paper not being published.

\end{document}
